---
title: "VAE_Metabolism"
author: "Christina Schmidt"
date: "23 May 2022"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE}
#load the needed libraries:
library(tidyverse) # used for data manipulation
library(rmarkdown) # used for paged_table function
```

# Samples
Here we use:\
1. The results of the VAE output of each SiRCle cluster (VAE-Results) for the different comparisons (young-old, Stage4-Stage1, PBRM1-BAP1).\
2. The results of the pathway analysis performing fGSEA on the SiRCle clusters.
3. The metabolic signatures used to perform the GSEA based on [Gaude et al](https://www.nature.com/articles/ncomms13041#Sec16) are used to select specific metabolic pathways of interest.\
4. mSigDB signatures: Biocarta, Hallmark, Reactome, KEGG.
```{r, warning=FALSE, message=FALSE}
#----------------------------------------------------------
#VAE Data
VAE_Stage <- read.csv("InputData/Input_VAE-Results/mean_Integrated_comparison_Stage IV-Stage I.csv", check.names=FALSE)
VAE_Age <- read.csv("InputData/Input_VAE-Results/mean_Integrated_comparison_old-young.csv", check.names=FALSE)
VAE_Mut <- read.csv("InputData/Input_VAE-Results/mean_Integrated_comparison_PBRM1-BAP1.csv", check.names=FALSE)

#----------------------------------------------------------
#fGSEA results
GSEA_Metab_Stage <-read.delim("InputData/Input_VAE-Results/fGSEA/Stage IV-Stage I_consolidated_metabolic_GSEA.tsv",check.names=FALSE)
GSEA_Path_Stage <-read.delim("InputData/Input_VAE-Results/fGSEA/Stage IV-Stage I_consolidated_pathways_GSEA.tsv",check.names=FALSE)
GSEA_Metab_Age <-read.delim("InputData/Input_VAE-Results/fGSEA/old-young_consolidated_metabolic_GSEA.tsv",check.names=FALSE)
GSEA_Path_Age <-read.delim("InputData/Input_VAE-Results/fGSEA/old-young_consolidated_pathways_GSEA.tsv",check.names=FALSE)
GSEA_Metab_Mut <-read.delim("InputData/Input_VAE-Results/fGSEA/PBRM1-BAP1_consolidated_metabolic_GSEA.tsv",check.names=FALSE)
GSEA_Path_Mut <-read.delim("InputData/Input_VAE-Results/fGSEA/PBRM1-BAP1_consolidated_pathways_GSEA.tsv",check.names=FALSE)

#----------------------------------------------------------
#Metabolic Pathway
Metabolic_Signature <-read.csv("InputData/Input_MetabolicPathways_Gaude/41467_2016_BFncomms13041_MOESM340_ESM.csv")

Correction_Metabolic_Signature <- read.csv("InputData/Input_MetabolicPathways_Gaude/41467_2016_BFncomms13041_MOESM341_ESM.csv")%>% 
    mutate(Unique = case_when(associated_Pathways =="1" ~ 'Unique',
                                  TRUE ~ 'In multiple Pathways'))
Metabolic_Signature <-merge(x=Metabolic_Signature, y=Correction_Metabolic_Signature, by.x ="gene", by.y="external_gene_name", all.x=TRUE)

# Other pathways
library(GSEABase)
gmt_to_csv <- function(gmtfile, fast = T){
  if(fast)
  {
    genesets = GSEABase::getGmt(con = gmtfile)
    genesets = unlist(genesets)
    
    gene_to_term =plyr::ldply(genesets,function(geneset){
      temp <- geneIds(geneset)
      temp2 <- setName(geneset)
      temp3 <- as.data.frame(cbind(temp,rep(temp2,length(temp))))
      
    },.progress = plyr::progress_text())
    names(gene_to_term) <- c("gene","term")
    return(gene_to_term[complete.cases(gene_to_term),])
  }
  else
  {
    genesets = getGmt(con = gmtfile)
    genesets = unlist(genesets)
    
    gene_to_term <- data.frame(NA,NA)
    names(gene_to_term) <- c("gene","term")
    for (geneset in genesets)
    {
      temp <- geneIds(geneset)
      temp2 <- setName(geneset)
      temp3 <- as.data.frame(cbind(temp,rep(temp2,length(temp))))
      names(temp3) <- c("gene","term")
      gene_to_term <- rbind(gene_to_term,temp3)
    }
    
    return(gene_to_term[complete.cases(gene_to_term),])
  }
}#Load the Pathways

KEGG <- gmt_to_csv("InputData/Input_MSigDB_Signatures/c2.cp.kegg.v6.2.symbols.gmt")
Reactome <- gmt_to_csv("InputData/Input_MSigDB_Signatures/c2.cp.reactome.v6.2.symbols.gmt")
Biocarta <- gmt_to_csv("InputData/Input_MSigDB_Signatures/c2.cp.biocarta.v6.2.symbols.gmt")
Hallmarks <- gmt_to_csv("InputData/Input_MSigDB_Signatures/h.all.v6.2.symbols.gmt")

Pathways <- rbind(KEGG, Reactome, Biocarta, Hallmarks)
```

# Plots VAE Mean Integrated Rank
This refers to the results of the VAE built for the individual SiRCle clusters.

## Lollipop Graphs
```{r, warning=FALSE}
#Prepare the DF needed:
InputLollipop <- function(Input, Condition1, Condition2){
  Cond1 <-Input[,-9]
  names(Cond1)[8]<-"Integrated Mean"
  Cond1$Condition <- paste(Condition1)
  
  Cond2 <-Input[,-8]
  names(Cond2)[8]<-"Integrated Mean"
  Cond2$Condition <- paste(Condition2)

  Graph <- rbind(Cond1,Cond2)
  Graph$`Integrated value_Round` <- round(Graph$`Integrated Mean`, digits = 2)
  Graph$Colour <- ifelse(Graph$Condition== paste(Condition1) , 3, 14)
  Output <- Graph
}

Graph_Stage<-InputLollipop(Input= VAE_Stage, 
                           Condition1="Stage 1", 
                           Condition2="Stage 4")

Graph_Age<-InputLollipop(Input= VAE_Age, 
                           Condition1="Young", 
                           Condition2="Old")

Graph_Mut<-InputLollipop(Input= VAE_Mut, 
                           Condition1="BAP1", 
                           Condition2="PBRM1")

#-----------------------------------------------
Lollipop_VAE <- function(Input, Signature, Signature_term, Comparison){
  SiRCle_Clusters <- c("MDE", "MDS", "TPDE", "TPDS", "TMDE", "TMDS", "MDS_TMDE","MDE_TMDS","TPDE_TMDS", "TPDS_TMDE")
    for (k in SiRCle_Clusters){
  Input_Cluster <- subset(Input, Regulation_Grouping_2 == paste(k))
      Pathway <- subset(Signature, term == paste(Signature_term))
      Plot  <- merge(x=Pathway,y=Input_Cluster, by.x="gene", by.y="external_gene_name", all.x=TRUE)
      Plot  <- na.omit(Plot)
  #Plot
   if(nrow(Plot)>1){    
  Dotplot1 <-ggplot(Plot, aes(x=reorder(gene, + `Integrated Mean`), y=`Integrated Mean`, label=`Integrated value_Round`)) + 
    geom_point(stat='identity', aes(size = `Integrated Mean`, col=Condition))  +
    geom_segment(aes(y =(Reduce(max,`Integrated Mean`)), 
                   x = gene, 
                   yend = `Integrated Mean`,
                   xend = gene), 
               color = "black") +
    scale_size(name="VAE Integrated Rank (Mean)",range = c(2,16))+
    geom_text(color="black", size=2) +
    labs(title=paste(Signature_term), 
        subtitle=paste(k)) + 
    ylim(((Reduce(min,Plot$`Integrated Mean`))-0.5),((Reduce(max,Plot$`Integrated Mean`))+0.5)) +
    theme_minimal() +
    coord_flip()+
    theme(plot.title = element_text(color = "black", size = 12, face = "bold"),
          plot.subtitle = element_text(color = "black", size=10),
          plot.caption = element_text(color = "black",size=9, face = "italic", hjust = 2.5))+
    labs(y="VAE integrated value", x="")
  ggsave(file=paste("Figures/Lollipop_IntegratedMean/", k,"_LollipopGraph_",Signature_term, Comparison,".pdf", sep=""), plot=Dotplot1, width=10, height=10)
  plot(Dotplot1)
   } else{
        print(paste(Signature_term, " pathway genes not present in SiRCle cluster ", k, sep=""))
        }
    }
}

#Run the function:
Pathway_Names <- Metabolic_Signature[!duplicated(Metabolic_Signature$term),]
Pathway_Names <- Pathway_Names$term

##Stage
for (i in Pathway_Names){
  Lollipop_VAE(Input=Graph_Stage,
            Signature= Metabolic_Signature,
            Signature_term=i,
            Comparison="_Stage IV-Stage I")
}

##Age
for (i in Pathway_Names){
  Lollipop_VAE(Input=Graph_Age,
            Signature= Metabolic_Signature,
            Signature_term=i,
            Comparison="_old-young")
}

##Mutation
for (i in Pathway_Names){
  Lollipop_VAE(Input=Graph_Mut,
            Signature= Metabolic_Signature,
            Signature_term=i,
            Comparison="_PBRM1-BAP1")
}
```
\
Next we plot all genes of different SiRCle clusters corresponding to the same metabolic pathway on one lollipop graph.\
```{r, warning=FALSE, message=FALSE}

Lollipop_All_VAE_Clusters <- function(Input, Signature, Signature_term, Comparison){
  #Select the pathway
      Pathway <- subset(Signature, term == paste(Signature_term))
      Plot  <- merge(x=Pathway,y=Input, by.x="gene", by.y="external_gene_name", all.x=TRUE)
      Plot  <- na.omit(Plot)
   #Plot
      if(nrow(Plot)>1){    
        Dotplot1 <-ggplot(Plot, aes(x=reorder(gene, + `Integrated Mean`), y=`Integrated Mean`, label=`Integrated value_Round`)) + 
    geom_point(stat='identity', aes(size = `Integrated Mean`, col=Condition))  +
    geom_segment(aes(y =(Reduce(max,`Integrated Mean`)), 
                   x = gene, 
                   yend = `Integrated Mean`,
                   xend = gene), 
               color = "black") +
   scale_size(name="VAE Integrated Rank (Mean)",range = c(2,16))+
    geom_text(color="black", size=2) +
    labs(title=paste(Signature_term), 
        subtitle="All SiRCle Clusters") + 
    ylim(((Reduce(min,Plot$`Integrated Mean`))-0.5),((Reduce(max,Plot$`Integrated Mean`))+0.5)) +
    theme_minimal() +
    coord_flip()+
    theme(axis.text.y = element_text(color = "black", size=9, face = "bold"),
          plot.title = element_text(color = "black", size = 12, face = "bold"),
          plot.subtitle = element_text(color = "black", size=10))+
          #plot.caption = element_text(color = "black",size=9, face = "italic", hjust = 2.5))+
    labs(y="VAE integrated value", x="")
  ggsave(file=paste("Figures/Lollipop_IntegratedMean/AllClusters_LollipopGraph_",Signature_term, Comparison,".pdf", sep=""), plot=Dotplot1, width=10, height=10)
  plot(Dotplot1)
     } else{
       print(paste(Signature_term, " pathway genes not present in any SiRCle cluster ", sep=""))
       }
}

#Run the function:
Pathway_Names <- Metabolic_Signature[!duplicated(Metabolic_Signature$term),]
Pathway_Names <- Pathway_Names$term

for (i in Pathway_Names){
  Lollipop_All_VAE_Clusters(Input=Graph_Stage,
            Signature= Metabolic_Signature,
            Signature_term=i,
            Comparison="_Stage IV-Stage I")
}

for (i in Pathway_Names){
  Lollipop_All_VAE_Clusters(Input=Graph_Age,
            Signature= Metabolic_Signature,
            Signature_term=i,
            Comparison="_old-young")
}

for (i in Pathway_Names){
  Lollipop_All_VAE_Clusters(Input=Graph_Mut,
            Signature= Metabolic_Signature,
            Signature_term=i,
            Comparison="_PBRM1-BAP1")
}

```

# Plots GSE
## fGSEA
This is based on the Gene set enrichment analysis (GSEA) results using the metabolic signatures and the pathways from the MSig DB (KEGG, Reactome, Biocarta, Hallmarks) . 
Here we use the following threshold:\
\
1. Given that the input data are quite small and specific (in the sense of pathways they correspond to). Indeed e.g. in MDE only 32 pathways that have a gene detected in them, which will change the results for multiple testing, hence we use p.val <0.25 as the significant threshold\
\
2. Given the pathway sizes are mostly really small (5 genes in one pathway), at least 10% of a pathway has to be detected.
```{r, warning=FALSE, message=FALSE}
# Add information on pathway size to the Signatures
SignatureCount <- function(Signature){
  DF<- Signature
  DF$Count <- 1
  DF$Count1 <- 1
  
  DF_Mean <- (DF[,3:4]) %>%
    group_by(DF$term) %>%
    summarise_all(funs(sum))%>%
    rename("DF$term"="term",
           "Count"="Genes_in_Pathway")
  DF_Mean <- DF_Mean[,-3]

  Output <- merge(x= DF[,1:2], y=DF_Mean, by="term", all.x=TRUE)
}


Metabolic_Signature_Count <- SignatureCount(Signature=Metabolic_Signature[,1:2])
Pathway_Count <- SignatureCount(Signature=Pathways)


# Add information on Pathway size to GSEA output and Select the Pathways to plot based on the threshold
Select_Pathway <- function(Input, Pathway_Count, FileName){
  DF <- merge(x= Input, y=Pathway_Count,by.x="Pathway", by.y="term", all.x=TRUE)
  DF$Percentage_of_Pathway_detected <-round(((DF$`fGSEA size`/DF$Genes_in_Pathway)*100),digits=2)
  DF  <- DF  %>%
    mutate(`In Threshold?` = case_when(`fGSEA pval` <= 0.25 &  Percentage_of_Pathway_detected>= 10 & Genes_in_Pathway >=5 ~ 'YES',
                                  `fGSEA pval` > 0.25 |  Percentage_of_Pathway_detected< 10 | Genes_in_Pathway < 5~ 'NO',
                                  TRUE ~ 'FALSE'))
  write.csv(DF,paste("OutputData/",FileName , ".csv"), row.names = FALSE)
  Output<- DF
}


GSEA_Metab_Stage <- Select_Pathway(Input=GSEA_Metab_Stage,
                                Pathway_Count=Metabolic_Signature_Count, 
                                FileName="Stage IV-Stage I_consolidated_metabolic_GSEA")
  
GSEA_Path_Stage <- Select_Pathway(Input=GSEA_Path_Stage,
                                Pathway_Count=Pathway_Count, 
                                FileName="Stage IV-Stage I_consolidated_pathways_GSEA")

GSEA_Metab_Age <-Select_Pathway(Input=GSEA_Metab_Age,
                                Pathway_Count=Metabolic_Signature_Count, 
                                FileName="old-young_consolidated_metabolic_GSEA")

GSEA_Path_Age <-Select_Pathway(Input=GSEA_Path_Age,
                                Pathway_Count=Pathway_Count, 
                                FileName="old-young_consolidated_pathways_GSEA")

GSEA_Metab_Mut <-Select_Pathway(Input=GSEA_Metab_Mut,
                                Pathway_Count=Metabolic_Signature_Count, 
                                FileName="PBRM1-BAP1_consolidated_metabolic_GSEA")

GSEA_Path_Mut <-Select_Pathway(Input=GSEA_Path_Mut,
                                Pathway_Count=Pathway_Count, 
                                FileName="PBRM1-BAP1_consolidated_pathways_GSEA")

```

```{r, warning=FALSE, message=FALSE}
#make the function:
Lollipop_fGSEA <- function(Input, fGSEA_Signature ,Comparison){
  #Make colour code
  keyvals<- ifelse(
    Input$Regulation_Grouping_2 == "MDS", "#CF4692",
    ifelse(Input$Regulation_Grouping_2 == "MDS&TMDE", "#A16BAA",
    ifelse(Input$Regulation_Grouping_2 == "MDE", "#6AAF43",
    ifelse(Input$Regulation_Grouping_2 == "MDE&TMDS", "#0E8D6D",
    ifelse(Input$Regulation_Grouping_2 == "TPDS", "#452D76",
    ifelse(Input$Regulation_Grouping_2 == "TPDS&TMDE", "#7E4090",
    ifelse(Input$Regulation_Grouping_2 == "TPDE", "#E58D26",
    ifelse(Input$Regulation_Grouping_2 == "TPDE&TMDS", "#844D13",
    ifelse(Input$Regulation_Grouping_2 == "TMDS", "#3A58A3",
    ifelse(Input$Regulation_Grouping_2 == "TMDE", "#E5322F",
    ifelse(Input$Regulation_Grouping_2 == "None", "grey",       
           "black")))))))))))
  names(keyvals)[is.na(keyvals)] <- "black"
  names(keyvals)[keyvals == 'black'] <- "NA"
  names(keyvals)[keyvals == '#CF4692'] <- "MDS"
  names(keyvals)[keyvals == '#A16BAA'] <- "MDS&TMDE"
  names(keyvals)[keyvals == '#6AAF43'] <- "MDE"
  names(keyvals)[keyvals == '#0E8D6D'] <- "MDE&TMDS"
  names(keyvals)[keyvals == '#452D76'] <- "TPDS"
  names(keyvals)[keyvals == '#7E4090'] <- "TPDS&TMDE"
  names(keyvals)[keyvals == '#E58D26'] <- "TPDE"
  names(keyvals)[keyvals == '#844D13'] <- "TPDE&TMDS"
  names(keyvals)[keyvals == '#3A58A3'] <- "TMDS"
  names(keyvals)[keyvals == '#E5322F'] <- "TMDE"
  names(keyvals)[keyvals == 'grey'] <- "None"
  
  #Select only pathways that meet the threshold
  Input_Select <-Input%>%
    subset(`In Threshold?`=="YES")
  
  if(nrow(Input_Select)>1){   
  #Plot
  Dotplot <-ggplot(Input_Select, aes(x=reorder(Pathway, +Percentage_of_Pathway_detected), y=`fGSEA NES`, label=Percentage_of_Pathway_detected)) + 
    geom_point(stat='identity', aes(size =Percentage_of_Pathway_detected, col=Regulation_Grouping_2)) +
     scale_color_manual(values = keyvals)+
    geom_segment(aes(y = 0, 
                   x = Pathway, 
                   yend = `fGSEA NES`,
                   xend = Pathway), 
               color = "black") +
   scale_size(name="Genes detected of pathway [%]",range = c(2,16))+
    geom_text(color="black", size=2) +
    labs(title= fGSEA_Signature,
        subtitle= Comparison,
        caption="*Threshold: p.val<0.25 & Genes detected of pathway > 10%") + 
    ylim(((Reduce(min,Input_Select$`fGSEA NES`))-0.5),((Reduce(max,Input_Select$`fGSEA NES`))+0.5)) +
    theme_bw() +
    coord_flip()+
    theme(plot.title = element_text(color = "black", size = 12, face = "bold"),
          plot.subtitle = element_text(color = "black", size=10),
          plot.caption = element_text(color = "black",size=9, face = "italic", hjust = 2.5))+
    labs(y="NES", x="")
  ggsave(file=paste("Figures/Lollipop_fGSEA/fGSEA_LollipopGraph",fGSEA_Signature, Comparison,".pdf", sep=""), plot=Dotplot, width=15, height=10)
  
  plot(Dotplot)
  } else{
       print(paste("GSEA of comparing ", Comparison, " patients does not lead to any altered pathways that pass the threshold criteria", sep=""))
       }
}

#Make the Graphs:
Lollipop_fGSEA(Input=GSEA_Metab_Stage,
               fGSEA_Signature= "_Metabolic_" ,
               Comparison="Stage IV-Stage I")

Lollipop_fGSEA(Input=GSEA_Path_Stage,
               fGSEA_Signature= "_Kegg-Reactome-Biocarta-hallmark_" ,
               Comparison="Stage IV-Stage I")

Lollipop_fGSEA(Input=GSEA_Metab_Age,
               fGSEA_Signature= "_Metabolic_" ,
               Comparison="old-young")

Lollipop_fGSEA(Input=GSEA_Path_Age,
               fGSEA_Signature= "_Kegg-Reactome-Biocarta-hallmark_" ,
               Comparison="old-young")

Lollipop_fGSEA(Input=GSEA_Metab_Mut,
               fGSEA_Signature= "_Metabolic_" ,
               Comparison="PBRM1-BAP1")

Lollipop_fGSEA(Input=GSEA_Path_Mut,
               fGSEA_Signature= "_Kegg-Reactome-Biocarta-hallmark_" ,
               Comparison="PBRM1-BAP1")

```

# Information about packages and versions used 
```{r}
sessionInfo()
```
